---
title: "Reproducible_Research Data Assessment"
author: "M.N.L"
date: "1/24/2021"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning =FALSE,
                      message = FALSE,
                      fig.width = 8,
                      fig.height= 4,
                      fig.align = "center",
                      fig.path ="Figs/"
)

```

## Activity Research Analysis
The response to the following questions are as follows

### Loading and Processing the data

```{r loading data and relevant packages}
library(readr);library(dplyr);library(lubridate);library("ggplot2")
library("Hmisc");library(reshape)
adata<-read_csv("/Users/mndoping/Desktop/DataScience/JHopkins Data Science/Reproducible Research/activity.csv")
head(adata)

```
### Average Number of Steps Taken per day

```{r total steps, echo=TRUE}
#Group data by date
adata_s<- adata %>% group_by(date) %>% 
summarise(Totalsteps=sum(steps,na.rm = TRUE))

#Plot of the histogram total steps per day
ggplot(adata_s, aes(x=Totalsteps))+geom_histogram(binwidth =5000, fill="lightcoral",
                                               color="black")

#Mean and median of total steps taken per day
adata_smean<-round(mean(adata_s$Totalsteps),2)
adata_smedian<-round(median(adata_s$Totalsteps),2)
```
The mean of the total steps per day is `r adata_smean` and the median of the total steps per day is `r adata_smedian`.

### Average daily Activity pattern

```{r daily activity pattern}
#Time series plot- Average daily pattern
adata_int<- adata %>% group_by(interval) %>% 
  summarise(steps_mean=mean(steps, na.rm = TRUE))
            
g2<-ggplot(adata_int, aes(x=interval, y=steps_mean))+geom_line(color="salmon2")+
  ggtitle("Time series plot of Average Daily Pattern ")
g2

#Determine 5min interval with maximum number of steps
max<-max(adata_int$steps_mean)
adata_max<-subset(adata_int, steps_mean==max)$interval
```
The 5-minute interval that contains maximum number of steps is `r adata_max`.

### Imputing Missing Values

```{r imputing}
library(knitr);library(kableExtra)
#Determining the total number of missing values data 
na_sum<-sum(is.na(adata))
c<-subset(adata, is.na(steps))
na_rows<- c %>% group_by(date) %>% 
  summarise(missing_sum=length(steps))
na_rows %>%
  kbl(caption = "Total missing rows per day") %>%
  kable_classic_2(full_width = F)

#Imputing mean using impute function
adata_imp<-impute(adata$steps,fun=mean)
adata$impute<-adata_imp

```
The total number of missing values is `r na_sum`. A break down of the missing values showed that these values were specific to entire days as seen above. The impute function from Hmisc package, using mean was used to replace missing values. 

```{r hist_impute}
#Plot histogram of total steps taken per day including imputed values
adata_new<- adata %>% group_by(date) %>% 
  summarise(Totalsteps=sum(impute))
g3<-ggplot(adata_new, aes(x=Totalsteps))+geom_histogram(binwidth =5000, fill="cyan1",color="black")

g3
```

```{r difference }
#Mean and median of total steps taken per day of new data with imputed values
adata_nmean<-round(mean(adata_new$Totalsteps),2)
adata_nmedian<-round(median(adata_new$Totalsteps),2)

#Impact of imputing values
mean_dif<-round((adata_nmean-adata_smean),2)
median_dif<-round((adata_nmedian-adata_smedian),2)
total_dif<-round((sum(adata_new$Totalsteps) - sum(adata_s$Totalsteps)),2)
```

The mean difference is `r mean_dif`, the median difference is `r median_dif`, and the total difference is `r total_dif`. Overall, as seen from values, imputed data shows higher values as expected.

### Difference in Activity Patterns Between Weekdays and Weekends

The difference in weekday and weekend activity patterns can be seen below
```{r difference patterns}
#Identifying Weekdays 
adata$wday<-wday(adata$date, label=TRUE)
for (i in 1:length(adata$wday)){
  if(adata$wday[i]=="Sat"|adata$wday[i]=="Sun"){
    adata$wdayL[i]<-1
  } else {adata$wdayL[i]<-2}
}

# Group  Data for plots
wkdy<-subset(adata, adata$wdayL==2)
wknd<-subset(adata, adata$wdayL==1)

wkdyi<- wkdy %>% group_by(interval) %>% 
  summarise(imp_wkdy=mean(impute))
wkndi<- wknd %>% group_by(interval) %>% 
  summarise(imp_wknd=mean(impute))
wkdyi$imp_wknd<-wkndi$imp_wknd
id<-data.frame("id"=c(1:288))
wkdyi<-cbind(id, wkdyi)

#Reshape Data for Plots
wkdyi<-melt(wkdyi, c("id","interval"))

g<-ggplot(wkdyi, aes(x=interval, y=value, color=variable))+geom_line()+
  facet_wrap(~variable, ncol=1)
g

```

